---
title: "Final Project - Covid-19 Analysis"
author: "Kesav Adithya Venkidusamy"
date: "11/16/2021"
output: pdf_document
---

# *Introduction*

  Coronavirus disease or COVID-19 is a global pandemic infectious disease caused by virus called sars-cov-2. Most people infected with the virus will experience mild to moderate respiratory illness and recover without requiring special treatment. However, some will become seriously ill and require medical attention. Older people and those with underlying medical conditions like cardiovascular disease, diabetes, chronic respiratory disease, or cancer are most likely to develop serious complications from COVID-19 illness. 
  
  The center for disease control and prevention (CDC) is the national public health agency of the United States. The agency’s main goal is the protection of public health and safety through control and prevention of disease, injury, and disability in US and worldwide. CDC plays an essential role in the response to COVID-19. The agency collects the data on regular basis and provide for public use. Among numerous datasets available in CDC, below are the ones considered for analysis
  
1.	Provisional COVID-19 deaths by sex and age
2.	Provisional COVID-19 deaths by week, sex and age
3.	Conditions contributing to COVID-19 deaths, by state and age, provisional 2020-21

  The COVID-19 pandemic also pushed many companies to develop new vaccines to minimize the severity of symptoms. These vaccines were developed rapidly and underwent clinical trials rigorous enough to meet FDA (Food and Drug Administration) requirement for emergency use. The government played a role in monitoring the adverse reactions of these newly developed vaccines with Vaccine Adverse Event Report System (VAERS). VARES is co-managed by the Central for Disease Control and Prevention (CDC) and U.S Food and Drug Administration (FDA).
  
  VARES accepts reports from people who have received vaccines and experienced adverse effects or from heathcare providers who are required by law to report:
  
1.	Any adverse event listed in the VARES table of reportable events following vaccination that occurs within the specified time period after vaccinations
2.	An adverse event listed by the vaccine manufacturer as a contradiction to further doses of vaccine

  VARES data is accessible by two mechanisms: by downloading raw data in comma-separated values (CSV) files for import into a database, spreadsheet or, by use of CDC WONDER online search tool. For this project, below datasets from VARES is considered.
  
1.	VARESDATA.csv
2.	VARESVAX.csv
3.	VARESSYMPTOMS.csv


# *The problem statement you addressed*

  The problem that I addressed is Covid-19 impact by age and people with underlying conditions. I analyzed if the deaths caused by Covid-19 virus is high for the older people having age greater than 55 compared to the young people whose age is less than 55. I also analyzed the impact of Covid-19 and deaths caused to the people with underlying condition like diabetes, blood pressure and stroke compared to the people who don't have any underlying conditions. Then, I analyzed Covid-19 vaccine data to see if it has any impact in controlling deaths.
  

# *How you addressed this problem statement*

I addressed the problem statement as follows:

  + Loading CDC (Covid-19 deaths) and VAERS (vaccines) data sets for analysis
  + Cleaning the data sets
  + Splitting and merging the data sets
  + Adding additional variables derived from existing variables to the data sets
  + Slicing and dicing the data sets
  + Viewing various metrics and graphs to perform the analysis
  

# *Analysis*

## *Loading the r libraries required for the analysis*

```{r lib}

library(readxl)
library(dplyr)
library(broom)
library(ggplot2)
library(GGally)
library(coefplot)
library(knitr)
library(tidyr)

```
  
## *Loading data sets for analysis*

### CDC Datasets
	Numerous Covid-19 related datasets are available for public use in CDC website. Those datasets feature case surveillance, deaths, population, race, sex and age. CDC reports aggregate counts of COVID-19 cases and death numbers daily and weekly online.
	
Provisional COVID-19 deaths by week, sex and age
  + Data as of – Date of Analysis
  + State - Jurisdiction of occurrence
  +	MMWR Week – MMWR week number
  +	End Week - Last week-ending date of data period
  +	Sex - Sex
  +	Age Group - Age group
  +	Total Deaths – Deaths from all causes of deaths
  +	COVID-19 Deaths - Deaths Involving COVID-19

Conditions contributing to COVID-19 deaths, by state and age, provisional 2020-21
  +	Start Date - First week-ending date of data period
  +	End Date - Last week-ending date of data period
  +	Group - Time-period Indicator for record: by Month, by Year, Total
  +	State - Jurisdiction of occurrence 
  +	Condition - Condition contributing to deaths involving COVID-19
  +	Age Group - Age group
  +	COVID-19 Deaths - COVID 19 Deaths

### VARES Dataset
  
  VARES data are distributed in three data sets, VARESVAX, VARESDATA and VARESSYMPTOMS. Data sets belong to year 2020 and 2021 will be used for this project. The code book for this data set is available in the below link

[Code Book](https://vaers.hhs.gov/docs/VAERSDataUseGuide_en_September2021.pdf)

```{r dataset}

covid19_week <- read.csv("Provisional_COVID-19_Deaths_by_Week_Sex_and_Age.csv")
covid19_condition <- read.csv("COVID-19_Deaths_by_State_and_Age.csv")
data20 <- read.csv("2020VAERSDATA.csv")
data21 <- read.csv("2021VAERSDATA.csv")
symptoms20 <- read.csv("2020VAERSSYMPTOMS.csv")
symptoms21 <- read.csv("2021VAERSSYMPTOMS.csv")
vaccine20 <- read.csv("2020VAERSVAX.csv")
vaccine21 <- read.csv("2021VAERSVAX.csv")

```

## *Cleaning the data sets*

```{r dataclean}

#Columns present in weekly covid-19 death data set
colnames(covid19_week)

print("Total number of rows and columns present in covid-19 weekly death data set")
dim(covid19_week)

#Columns present in covid-19 death occurred due to underlying condition 
colnames(covid19_condition)

print("Total number of rows and columns present in covid-19 death data set with underlying conditon")
dim(covid19_condition)

#Applying filter to the data sets as defined above
covid19_week_filter <- covid19_week %>% filter(State == "United States" & Sex == "All Sex" & Age.Group != "All Ages") 
dim(covid19_week_filter)

covid19_cond_filter <- covid19_condition %>% filter(State == "United States" & Group == "By Month" & Age.Group != "All Ages" & Age.Group != "Not stated")
dim(covid19_cond_filter)

#Removing unwanted columns that are not required for the analysis
covid19_week_cols <- c(2,4,5,6,8)
covid19_week_final <- covid19_week_filter[,covid19_week_cols]
colnames(covid19_week_final)

covid19_condition_cols <- c(2,3,5,6,7,8,11,12)
covid19_condition_final <- covid19_cond_filter[,covid19_condition_cols]
colnames(covid19_condition_final)

#Merge data sets by year for VARES
merged_vaccine_20 <- merge(data20, symptoms20)
merged_vaccine_20 <- merge(merged_vaccine_20, vaccine20)
dim(merged_vaccine_20)
colnames(merged_vaccine_20)

merged_vaccine_21 <- merge(data21, symptoms21)
merged_vaccine_21 <- merge(merged_vaccine_21, vaccine21)
dim(merged_vaccine_21)
colnames(merged_vaccine_21)

#Cleaning VARES data set. From the entire data set, We have to choose vaccines given for COVID-19 only. So, Applying the filter to filter covid-19 records

filter_vaccine_20 <- filter(merged_vaccine_20, grepl("COVID19", merged_vaccine_20$VAX_TYPE))
filter_vaccine_21 <- filter(merged_vaccine_21, grepl("COVID19", merged_vaccine_21$VAX_TYPE))

#Removing unwanted columns from the data set
vaccine_cols <- c(1,3,4,7,9,10,12,21,23,28,35,36,38,40,42,44,46,47,48,49,52)

vaccine_20_final <- filter_vaccine_20[,vaccine_cols]
vaccine_21_final <- filter_vaccine_21[,vaccine_cols]

#The columns present in vaccine 2020 data set
colnames(vaccine_20_final)

#Total number of rows and columns present in the data set
dim(vaccine_20_final)

#The columns present in vaccine 2021 data set
colnames(vaccine_21_final)

#Total number of rows and columns present in the data set
dim(vaccine_21_final)

```

## *Final data sets*

```{r finaldata}

#The final data sets after cleaning and before slicing and dicing
#covid-19 weekly death count by Age
print(str(covid19_week_final))

#Covid-19 monthly deaths by age with underlying condition
print(str(covid19_condition_final))

#Covid-19 Vaccine data for 2020 and 2021
print(str(vaccine_20_final))
print(str(vaccine_21_final))

```

## *Adding additional variable to final data sets*

I will be adding a variable called people to covid19_weekly data set which tells if the people is young or old based on the age. In addition, I will be adding a variable called "condition_flag" to covid19_condition data set which tells if the people had underlying conditions.

```{r enhancedataset}

old <- c("55-64 Years","65-74 Years","75-84 Years","85 Years and Over")
covid19_week_final$people <- ifelse(covid19_week_final$Age.Group %in% old, "Old", "Young") 
colnames(covid19_week_final)
print(str(covid19_week_final))

covid19_condition_final$condition_flag <- ifelse(covid19_condition_final$Condition.Group == "COVID-19","No","Yes")
print(str(covid19_condition_final))

```

## *Different ways to view the data*

Some of the different ways to look at the data set

* Provisional COVID-19 deaths by week, sex and age
  + Age
  + Week
  + Covid-19 Deaths
  + People (derived variable based on age of the people)
  
* Conditions contributing to COVID-19 deaths, by state and age, provisional 2020-21
  + Age
  + Condition.Group 
  + Covid-19 Deaths
  + condition_flag (derived variable based on condition.group)
  
* Vaccine data sets
  + Age
  + Died
  + VAERS_ID
  
## *Slicing and Dicing the data sets*

### Covid-19 Weekly death data set

```{r analysis}

covid19_week_final$COVID.19.Deaths <- as.numeric(covid19_week_final$COVID.19.Deaths)
print(str(covid19_week_final,10))

#Total deaths by Covid-19 for Young and old People
covid19_week_final %>% group_by(people) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

#Slicing the data set based on People (Young and Old)
covid19_week_young <- filter(covid19_week_final, people=="Young" & COVID.19.Deaths>0)
dim(covid19_week_young)

covid19_week_old <- filter(covid19_week_final, people=="Old" & COVID.19.Deaths>0)
dim(covid19_week_old)

#Printing the total number of deaths for young and old people
cat("Total number of covid-19 deaths for young people: ",sum(covid19_week_young$COVID.19.Deaths))
cat("Total number of covid-19 deaths for old people: ",sum(covid19_week_old$COVID.19.Deaths))

#Histograms on covid-19 death for young and old people
ggplot(covid19_week_young, aes(COVID.19.Deaths)) + geom_histogram(bins=30)
ggplot(covid19_week_old, aes(COVID.19.Deaths)) + geom_histogram(bins=30)

#Summary of weekly covid-19 deaths data set
summary(covid19_week_young)
summary(covid19_week_old)

cat("The variance of death count for young people: ", var(covid19_week_young$COVID.19.Deaths))
cat("The standard deviation of death count for young people: ", sd(covid19_week_young$COVID.19.Deaths))

cat("The variance of death count for old people: ", var(covid19_week_old$COVID.19.Deaths))
cat("The standard deviation of death count for old people: ", sd(covid19_week_old$COVID.19.Deaths))

covid19_deaths_by_people <- covid19_week_final %>% group_by(Age.Group) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

#Remove Years from the Age.Group
covid19_deaths_by_people$Age.Group <- gsub("Years","", covid19_deaths_by_people$Age.Group)

covid19_deaths_by_people

ggplot(covid19_deaths_by_people, aes(x=Age.Group, y=COVID19_Deaths)) + geom_bar(stat = "identity")

```

### Covid-19 death underlying condition

```{r condition}

#Converting datatype to numeric 
covid19_condition_final$COVID.19.Deaths <- as.numeric(covid19_condition_final$COVID.19.Deaths)
print(str(covid19_condition_final))

#Slicing the data set based on People with and without condition
covid19_condition_no <- filter(covid19_condition_final, condition_flag=="No" & COVID.19.Deaths>0)
dim(covid19_condition_no)

covid19_condition_yes <- filter(covid19_condition_final, condition_flag=="Yes" & COVID.19.Deaths>0)
dim(covid19_condition_yes)

#Printing the total number of deaths for young and old people
cat("Total number of covid-19 deaths for the people without underlying condition: ",sum(covid19_condition_no$COVID.19.Deaths))
cat("Total number of covid-19 deaths for the people with underlying condition: ",sum(covid19_condition_yes$COVID.19.Deaths))

#Histograms on covid-19 death for young and old people
ggplot(covid19_condition_no, aes(COVID.19.Deaths)) + geom_histogram(bins=30)
ggplot(covid19_condition_yes, aes(COVID.19.Deaths)) + geom_histogram(bins=30)

#Summary of covid-19 deaths
summary(covid19_condition_no)
summary(covid19_condition_yes)

cat("The variance of death count for the people without underlying condition: ", var(covid19_condition_no$COVID.19.Deaths))
cat("The standard deviation of death count for the people without underlying condition: ", sd(covid19_condition_no$COVID.19.Deaths))

cat("The variance of death count for the people with underlying condition: ", var(covid19_condition_yes$COVID.19.Deaths))
cat("The standard deviation of death count for the people with underlying condition: ", sd(covid19_condition_yes$COVID.19.Deaths))

head(covid19_condition_final)

covid19_condition_final$COVID.19.Deaths <- as.numeric(covid19_condition_final$COVID.19.Deaths)

covid19_deaths_by_condition <- covid19_condition_final %>% group_by(condition_flag,Age.Group) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

covid19_deaths_by_condition

covid19_deaths_by_condition.long <- covid19_deaths_by_condition %>% gather("Stat","Value",-Age.Group)

covid19_deaths_by_condition

ggplot(covid19_deaths_by_condition, aes(x=factor(Age.Group), y=COVID19_Deaths, fill = condition_flag)) + geom_bar(stat = "identity", width = 0.4, position = "dodge")

```

### Vaccine data sets

```{r vaccine}

#Combining vaccine data for both 2020 and 2021
vaccine_df <- union(vaccine_20_final, vaccine_21_final)

#Age Analysis
age_variable <- vaccine_df[-is.na(vaccine_df$AGE_YRS)]
age_hist <- ggplot(vaccine_df, aes(AGE_YRS)) + geom_histogram(bins=30)
age_hist

#Death data analysis

vaccine_died <- dplyr::filter(vaccine_df, grepl("Y",vaccine_df$DIED))
colnames(vaccine_died)
vaccine_died_nodup <- vaccine_died |> dplyr::distinct(VAERS_ID, .keep_all = TRUE) 
dim(vaccine_died_nodup)

cat("Total number of people died after taking vaccine: ", length(unique(vaccine_died$VAERS_ID)))

#Splitting the data set into young and old based on age. 
vaccine_died_young <- filter(vaccine_died, AGE_YRS<55)
dim(vaccine_died_young)
cat("Total number of young people died after taking vaccine: ", length(unique(vaccine_died_young$VAERS_ID)))

vaccine_died_old <- filter(vaccine_died, AGE_YRS>=55)
dim(vaccine_died_old)
cat("Total number of old people died after taking vaccine: ", length(unique(vaccine_died_old$VAERS_ID)))

```

# *Implications*

## CDC weekly death count data set

  From this data set, I could see the death count of young people (age < 55) is less compared to old people (age >= 55). The death count during the initial months were less as Covid-19 infection started spreading and peaked in the later months on 2020 and initial months of 2021, and again started going down from middle of 2021 due to vaccinations.

  The histogram for the Covid-19 deaths for young people is positively skewed distribution whereas the histogram for the Covid-19 deaths for old people is also positively skewed distribution but shows some pattern for multiple distribution as well.

  Bar chart also depicts the same where the count of covid-19 deaths is higher for old people compared to young people. 

## CDC's Covid-19 death underlying condition data set

  This data set tells that the death count of the people without any underlying condition is less compared to those people with underlying condition. 

  The histograms for the Covid-19 deaths for the people with and without underlying conditions are positively skewed distribution. This is because the covid-19 death count is high during 2020 and 1st quarter of 2021. From 2nd quarter of 2021, the count started decreasing.
  
## VAERS Vaccine data set 

  The vaccine data set also shows the death count of the people having young age (less than 54) is less compared to the old people having age greater than 55. 

## Average death due to covid-19 for young and old people

* Average covid-19 death count for young people: 109.6
* Average covid-19 death count for old people: 514.4

## Variance and standard deviation of covid-19 death for young and old people

* The variance and standard deviation of covid-19 death count for young people: 30973.93 and 175.9941
* The variance and standard deviation of covid-19 death count for old people: 99185.96 and 314.938

## Average death due to covid-19 for the people with and without underlying condition 

* Average covid-19 death count for the people without underlying condition: 164
* Average covid-19 death count for the people with underlying condition: 210

## Variance and standard deviation of covid-19 death for the people with and without underlying condition?

* The variance and standard deviation of covid-19 death for the people without underlying condition: 68582.8 and 261.8832
* The variance and standard deviation of covid-19 death for the people with underlying condition: 61526.05 and 248.0445

## Role age played in covid-19 deaths

```{r summary}

#Calculate total deaths by age
cat("Number of covid-19 deaths by age group:\n")
covid19_week_final %>% group_by(Age.Group) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

#Filtering the data till Aug 2021 and applying group by to calculate total deaths by end week
cat("Number of covid-19 deaths by week:\n")
covid19_week_final %>% filter(as.Date(End.Week, format= "%m/%d/%Y") < "2021-09-01") %>% group_by(as.Date(End.Week, format= "%m/%d/%Y")) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

```

## Role underlying condition played in covid-19 deaths

```{r under}

#Total deaths by Covid-19 for the people with and without underlying condition
cat("Number of deaths by underlying condition: \n")
covid19_condition_final %>% group_by(condition_flag) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

#Death count by underlying condition
cat("Number of covid-19 deaths by underlying condition")
covid19_condition_yes %>% group_by(Condition.Group) %>% summarise(COVID19_Deaths=sum(COVID.19.Deaths, na.rm = TRUE))

```

## Number of covid-19 deaths after taking vaccine by age and manufacture

```{r vaccine_analysis}
cat("Number of covid-19 deaths after taking vaccination by age")
death_age <- table(vaccine_died$AGE_YRS)
print(death_age)


cat("Number of covid-19 deaths after taking vaccination by manufacture")
death_type <- table(vaccine_died$VAX_MANU)
print(death_type)

cat("Number of covid-19 deaths after taking vaccination by state")
death_state <- table(vaccine_died$STATE)
print(death_state)

```

# *Limitations*

Some of the limitations are below

  + The number of deaths reported by CDC may not be accurate. Only the deaths occurred in hospital certified by doctors are reported in the data set. 
  + Number of reports may increase in response to media attention
  + I would want to analyze the percentage of vaccines given across the states and check for correlations between number of vaccine and deaths. However, I am unsure of how far I will get due to limitations in data.
  + Moreover, as part of initial phase, Covid-19 vaccines are given only to the people who are 18 years older. So, the vaccine data sets used is not complete.
  + The analysis on Covid-19 deaths by age and underlying condition has been done for Unites  States as a whole. The same can be extended to state level analysis but not done due to limitation with data set.
  + It is generally not possible to find out deaths from VARES data if a vaccine caused the adverse effect. 
  
# *Concluding Remarks*

  Based on analysis of the data sets extracted from CDC and VAERS, I conclude that adverse events caused by Covid-19 is high for the old people having age greater than 55 years compared to young people whose age is less than 55. In addition, the adverse effect caused by Covid-19 is high for the people having underlying condition compared to those who are healthy. The metrics and graphs generated out of this data sets also proving similar information. However, we need to keep in mind that the total number of death reported by CDC may not have complete information.
  